Structured Versus Decorated Cospans

### 2020-12-01 cospans.tex
eliminated stuff about braided pseudomonoids
finished section 5.4, on open Petri nets

\begin{thm}
Given two finitely cocomplete categories $\A$ and $\A'$, two symmetric lax monoidal pseudofunctors $F \maps \A \to \Cat$ and $F' \maps \A' \to \Cat$, a finite colimit preserving functor $H \maps \A \to \A'$, a symmetric lax monoidal pseudofunctor $E \maps \Cat \to \Cat$ and a 2-isomorphism $\theta \maps EF \Rightarrow F' H$ as in the following diagram,
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$\A$};
\node (B) at (1,0) {$\Cat$};
\node (C) at (0,-1) {$\mathsf{A'}$};
\node (C) at (0,-1) {$\A'$};
\node (D) at (1,-1) {$\Cat$};
\node (E) at (0.5,-0.5) {$\Swarrow \theta$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$F$} (B)
(A) edge node[left]{$H$} (C)
(B) edge node[right]{$E$} (D)
(C) edge node[above]{$F'$} (D);
\end{tikzpicture}
\] 
the triple $(H,E,\theta)$ induces a symmetric monoidal double functor $\lH \maps F\lCsp \to F'\lCsp$ as defined above.
\end{thm}
\begin{proof}  See \cite[Theorem 4.2.1]{CourserThesis}. \end{proof}

@@

Since both double categories are symmetric monoidal under the running assumptions (\cref{DC,thm:SC}), we will now show that this double isomorphism is a symmetric monoidal one, namely $\lE=(\lE_0,\lE_1)$ is a symmetric strong monoidal double functor as per \cref{defn:monoidal_double_functor}. Indeed, $\lE_0=\mathrm{id}_\A$ is a braided monoidal functor trivially. For $\lE_1$, given two structured cospans $M=(a,\bot_a)\to(m,x)\leftarrow(b,\bot_b)$ and $M'=(a',\bot_{a'})\to(m',x')\leftarrow(b',\bot_{b'})$ in $_L\lCsp(\inta F)$, we compute the appropriate decorated cospans
\begin{align*}
\lE(M)\ot\lE(M')&\;\stackrel{\cref{eq:tensordecoration}}{=}\;a+a'\xrightarrow{i+i'}m+m'\xleftarrow{o+o'}b+b' \textrm{ with }\one\xrightarrow{x\times x'}F(m)\times F(m')\xrightarrow{\phi_{m,m'}}F(m+m') \\
\lE(M\ot M')&\;\stackrel{\cref{eq:tensorstuctured}}{=}\;\lE\Big((a+a',\bot_{a+a'})\to\underbrace{(m+m',F(\iota_m)x+F(\iota_{m'})x')}_{\cref{eq:fibrewisecoprod}}\leftarrow(b+b',\bot_{b+b'})\Big)%=a+a'\xrightarrow{i+i'}m+m'\xleftarrow{o+o'}b+b' \textrm{ with }\phi_{m,m'}(x,x')
\end{align*}
which are uniquely isomorphic $F(\iota_m)x+F(\iota_{m'})\cong \phi_{m,m'}(x,x')$ in an analogous way to \cref{eq:isomorphismdiag}, as also discussed after \cref{cor:fcocMonGroth}.
Moreover, the monoidal unit for the category of arrows of $F\lCsp$ is $0_\A\xrightarrow{\raisebox{-3pt}[0pt][0pt]{$=$}}0_\A\xleftarrow{\raisebox{-3pt}[0pt][0pt]{$=$}}0_\A$ with the trivial decoration $\phi_0\maps \one\to F(0_A)$,
which is an initial object and as such isomorphic to $0_{F(0)}$
%=\lE((0_A,)\xrightarrow{\raisebox{-3pt}[0pt][0pt]{$=$}}(0_A,\phi_0)\xleftarrow{\raisebox{-3pt}[0pt][0pt]{$=$}}(0_A,\phi_0))$
again by \cref{eq:explicitstructure2} and the discussion after \cref{cor:fcocMonGroth}.
It can be verified that with the above structure, $\lE_1$ is also a braided monoidal functor, and the rest of the axioms of \cref{defn:monoidal_double_functor} hold.
\end{proof}

@@

Decorated cospans were introduced by Fong \cite{Fong,FongThesis} to describe open systems as cospans equipped with extra data; they were then applied to open electrical circuits \cite{BF}, Markov processes \cite{BFP}, and chemical reaction networks \cite{BP}.  Unfortunately, some of these applications were marred by technical flaws, which were later fixed using structured cospans \cite[Sec.\ 5]{BC}.  Our new approach to decorated cospans provides another solution to these problems.  Here we compare the two approaches in applications to graphs, electrical circuits, Petri nets, reaction networks and dynamical systems.   In many cases \cref{thm:equiv} shows that the structured and decorated cospan approaches are equivalent, but we show that open dynamical systems can only be treated using decorated cospans.

@@

Starting from the symmetric lax monoidal functor $F \maps (\Fin\Set, +) \to (\Cat,\times)$, \cref{DC} gives us a symmetric monoidal double category $F\lCsp$ in which:
\begin{itemize}
\item objects are sets,
\item a vertical 1-morphism from $X$ to $X'$ is a function $f \maps X \to X'$,
\item a horizontal 1-cell from $X$ to $Y$ is a pair
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$X$};
\node (B) at (1,0) {$N$};
\node (C) at (2,0) {$Y$};
\node (D) at (3.25,0) {$G \in F(N)$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$i$} (B)
(C) edge node[above]{$o$} (B);
\end{tikzpicture}
\]
which can also be thought of as an open graph from $X$ to $Y$,
\item a 2-morphism
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$X$};
\node (B) at (1,0) {$N$};
\node (C) at (2,0) {$Y$};
\node (A') at (0,-1) {$X'$};
\node (C') at (2,-1) {$Y'$};
\node (D) at (1,-1) {$N'$};
\node (E) at (3,0) {$G \in F(N)$};
\node (F) at (3,-1) {$G' \in F(N')$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$i$} (B)
(C) edge node[above]{$o$} (B)
(A) edge node[left]{$f$} (A')
(C) edge node[right]{$g$} (C')
(C') edge node [above] {$o'$} (D)
(A') edge node [above] {$i'$} (D)
(B) edge node [left] {$h$} (D);
\end{tikzpicture}
\]
is a commuting diagram in $\Fin\Set$ together with a morphism $\tau \maps F(h)(G) \to G'$ in $F(N')$.
\end{itemize}

@@

We can also describe $\La$-circuits using decorated cospans.   There is a symmetric lax monoidal functor $F \maps (\Fin\Set, +) \to (\Cat, \times)$ such that for any finite set $N$, the category $F(N)$ has:
\begin{itemize}
\item objects being $\La$-\define{graph structures} on $N$: that is, $\La$-graphs where the set of nodes is $N$,
\item morphisms are morphisms of $\La$-graphs that are the identity on the set of nodes.
\end{itemize}
This gives a symmetric monoidal double category $F \lCsp$, and using \cref{thm:equiv} we can show that this is equivalent, as a symmetric monoidal double category, to ${}_L \lCsp(\Graph_\La)$.

@@

There is a category $\Petri$ with Petri nets as objects, where a morphism from the Petri net 
$s, t \maps T \to \N[S]$ to the Petri net $s', t' \maps T' \to \N[S']$ is a pair of functions $f \maps T \to T', g \maps S \to S'$ such that the following diagrams commute:
	\xymatrix{ 
		T \ar[d]_f  \ar[r]^-{s} & \N[S] \ar[d]^-{\N[g]} \\	
		T' \ar[r]^-{s'} & \N[S'] 
	\qquad
	\xymatrix{ 
		T \ar[d]_f  \ar[r]^-{t} & \N[S] \ar[d]^-{\N[g]} \\	
		T' \ar[r]^-{t'} & \N[S'] . 
There is a functor $R \maps \Petri \to \Set$ sending any Petri net to its set of places, and this has
a left adjoint $L \maps \Set \to \Petri$ sending any set $S$ to the Petri net with $S$ as its set of species and no transitions \cite[Lemma 11]{BM}.   Since both $\Set$ and $\Petri$ have finite colimits and $L$ preserves them, Theorem \cref{thm:SC} yields a symmetric monoidal double category $\lOpen(\Petri)$ of open Petri nets which has:
\item objects given by sets,
\item vertical 1-morphisms given by functions,
\item horizontal 1-cells as \define{open Petri nets}, which are cospans in $\Petri$ of the form:
\node (D) at (-3,0) {$L(X)$};
\node (E) at (-2,0) {$P$};
\node (F) at (-1,0) {$L(Y)$};
(D) edge node [above] {$i$} (E)
(F) edge node [above] {$o$} (E);
\item 2-morphisms as maps of cospans in $\Petri$ of the form:
\node (E) at (3,0) {$L(X_1)$};
\node (F) at (5,0) {$L(Y_1)$};
\node (G) at (4,0) {$P_1$};
\node (E') at (3,-1) {$L(X_2)$};
\node (F') at (5,-1) {$L(Y_2)$.};
\node (G') at (4,-1) {$P_2$};
(F) edge node[above]{$O_1$} (G)
(E) edge node[left]{$L(f)$} (E')
(F) edge node[right]{$L(g)$} (F')
(G) edge node[left]{$\alpha$} (G')
(E) edge node[above]{$I_1$} (G)
(E') edge node[above]{$I_2$} (G')
(F') edge node[above]{$O_2$} (G');
We can also obtain a similar double category using decorated cospans: define a pseudofunctor $F \maps \Set \to \Cat$ where given a set $s$, $F(s)$ is the category of all Petri net structures with $s$ as its set of species. This pseudofunctor $F$ is symmetric lax monoidal as both $(\Set,+,\emptyset)$ and $(\Cat,\times,1)$ are symmetric monoidal and given Petri nets $P \in F(s)$ and $P' \in F(s')$, we can place them side by side and consider them together as a single Petri net $P+P' \in F(s+s')$ with set of species $s+s'$, and thus we have natural transformations $\phi_{s,s'} \maps F(s) \times F(s') \to F(s+s')$ for any two sets $s$ and $s'$. The other structure morphism between monoidal units $\phi \maps 1_{\Petri} \to F(\emptyset)$ is defined by the unique morphism from the empty Petri net with the empty set for its set of species to the only possible Petri net on the empty set, which is also the empty Petri net. All of the diagrams that are required to commute are straightforward. Using \cref{DC}, we obtain a symmetric monoidal double category $F\lCsp$ which has:
\item{objects given by sets,}
\item{vertical 1-morphisms given by functions,}
\item{horizontal 1-cells given by pairs:
\node (D) at (-3,0) {$X$};
\node (E) at (-2,0) {$Z$};
\node (F) at (-1,0) {$Y$};
\node (A) at (0,0) {$P \in F(Z)$};
(D) edge node [above] {$i$} (E)
(F) edge node [above] {$o$} (E);
\item{2-morphisms as maps of cospans in $\Set$:
\node (E) at (3,0) {$X_1$};
\node (F) at (5,0) {$Y_1$};
\node (G) at (4,0) {$Z_1$};
\node (E') at (3,-1) {$X_2$};
\node (F') at (5,-1) {$Y_2$};
\node (G') at (4,-1) {$Z_2$};
\node (A) at (6,0) {$P_1 \in F(Z_1)$};
\node (B) at (6,-1) {$P_2 \in F(Z_2)$};
(F) edge node[above]{$o_1$} (G)
(E) edge node[left]{$f$} (E')
(F) edge node[right]{$g$} (F')
(G) edge node[left]{$h$} (G')
(E) edge node[above]{$i_1$} (G)
(E') edge node[above]{$i_2$} (G')
(F') edge node[above]{$o_2$} (G');
together with a morphism of Petri nets $\tau \maps F(h)(P_1) \to P_2$ in $F(Z_2)$.}
Thus we have a symmetric monoidal double category $\lOpen(\Petri)$ of open Petri nets obtained from structured cospans and a symmetric monoidal double category $F\lCsp$ of open Petri nets obtain from decorated cospans, and of course, we have an equivalence $\lOpen(\Petri) \simeq F\lCsp$ of symmetric monoidal double categories by \cref{thm:equiv}.

%%%DISCUSS OPEN COMMUTATIVE MONOIDAL CATEGORIES!!!

@@

Let $\Dbl$ denote the 2-category of double categories, double functors and double transformations. One can check that $\Dbl$ has finite products, and in any 2-category with finite products we can define a `pseudomonoid', which is a categorified analogue of a monoid \cite{DS}. For example, a pseudomonoid in $\Cat$ is a monoidal category.   We can also define braided and symmetric pseudomonoids, which in $\Cat$ are braided and symmetric monoidal categories.

>>>>

\begin{thm}
\label{thm:functoriality}
Given two finitely cocomplete categories $\A$ and $\A'$, two symmetric lax monoidal pseudofunctors $F \maps \A \to \Cat$ and $F' \maps \A' \to \Cat$, a finite colimit preserving functor $H \maps \A \to \A'$, a symmetric lax monoidal pseudofunctor $E \maps \Cat \to \Cat$ and a 2-isomorphism $\theta \maps EF \Rightarrow F' H$ as in the following diagram,
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$\A$};
\node (B) at (1,0) {$\Cat$};
\node (C) at (0,-1) {$\mathsf{A'}$};
\node (C) at (0,-1) {$\A'$};
\node (D) at (1,-1) {$\Cat$};
\node (E) at (0.5,-0.5) {$\Swarrow \theta$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$F$} (B)
(A) edge node[left]{$H$} (C)
(B) edge node[right]{$E$} (D)
(C) edge node[above]{$F'$} (D);
\end{tikzpicture}
\] 
the triple $(H,E,\theta)$ induces a symmetric monoidal double functor $\lH \maps F\lCsp \to F'\lCsp$ as defined above.
\end{thm}

\begin{proof}  See \cite[Theorem 4.2.1]{CourserThesis}. \end{proof}

@@

Since both double categories are symmetric monoidal under the running assumptions (\cref{DC,thm:SC}), we will now show that this double isomorphism is a symmetric monoidal one, namely $\lE=(\lE_0,\lE_1)$ is a symmetric strong monoidal double functor as per \cref{defn:monoidal_double_functor}. Indeed, $\lE_0=\mathrm{id}_\A$ is a braided monoidal functor trivially. For $\lE_1$, given two structured cospans $M=(a,\bot_a)\to(m,x)\leftarrow(b,\bot_b)$ and $M'=(a',\bot_{a'})\to(m',x')\leftarrow(b',\bot_{b'})$ in $_L\lCsp(\inta F)$, we compute the appropriate decorated cospans
\begin{align*}
\lE(M)\ot\lE(M')&\;\stackrel{\cref{eq:tensordecoration}}{=}\;a+a'\xrightarrow{i+i'}m+m'\xleftarrow{o+o'}b+b' \textrm{ with }\one\xrightarrow{x\times x'}F(m)\times F(m')\xrightarrow{\phi_{m,m'}}F(m+m') \\
\lE(M\ot M')&\;\stackrel{\cref{eq:tensorstuctured}}{=}\;\lE\Big((a+a',\bot_{a+a'})\to\underbrace{(m+m',F(\iota_m)x+F(\iota_{m'})x')}_{\cref{eq:fibrewisecoprod}}\leftarrow(b+b',\bot_{b+b'})\Big)%=a+a'\xrightarrow{i+i'}m+m'\xleftarrow{o+o'}b+b' \textrm{ with }\phi_{m,m'}(x,x')
\end{align*}
which are uniquely isomorphic $F(\iota_m)x+F(\iota_{m'})\cong \phi_{m,m'}(x,x')$ in an analogous way to \cref{eq:isomorphismdiag}, as also discussed after \cref{cor:fcocMonGroth}.
Moreover, the monoidal unit for the category of arrows of $F\lCsp$ is $0_\A\xrightarrow{\raisebox{-3pt}[0pt][0pt]{$=$}}0_\A\xleftarrow{\raisebox{-3pt}[0pt][0pt]{$=$}}0_\A$ with the trivial decoration $\phi_0\maps \one\to F(0_A)$,
which is an initial object and as such isomorphic to $0_{F(0)}$
%=\lE((0_A,)\xrightarrow{\raisebox{-3pt}[0pt][0pt]{$=$}}(0_A,\phi_0)\xleftarrow{\raisebox{-3pt}[0pt][0pt]{$=$}}(0_A,\phi_0))$
again by \cref{eq:explicitstructure2} and the discussion after \cref{cor:fcocMonGroth}.
It can be verified that with the above structure, $\lE_1$ is also a symmetric monoidal functor, and the rest of the axioms of \cref{defn:monoidal_double_functor} hold.
\end{proof}

@@

Decorated cospans were introduced by Fong \cite{Fong,FongThesis} to describe open systems as cospans equipped with extra data.  They were then applied to open electrical circuits \cite{BF}, Markov processes \cite{BFP}, and chemical reaction networks \cite{BP}.  Unfortunately, some of these applications were marred by technical flaws, which were later fixed using structured cospans \cite[Sec.\ 5]{BC}.  Our new approach to decorated cospans provides another solution to these problems.  Here we compare the two approaches in applications to graphs, electrical circuits, Petri nets, reaction networks and dynamical systems.   In many cases \cref{thm:equiv} shows that the structured and decorated cospan approaches are equivalent, but we show that open dynamical systems can only be treated using decorated cospans.

@@

Starting from the symmetric lax monoidal pseudofunctor $F \maps (\Fin\Set, +) \to (\Cat,\times)$, \cref{DC} gives us a symmetric monoidal double category $F\lCsp$ in which:
\begin{itemize}
\item objects are sets,
\item a vertical 1-morphism from $X$ to $X'$ is a function $f \maps X \to X'$,
\item a horizontal 1-cell from $X$ to $Y$ is a pair
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$X$};
\node (B) at (1,0) {$N$};
\node (C) at (2,0) {$Y$};
\node (D) at (3.25,0) {$G \in F(N)$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$i$} (B)
(C) edge node[above]{$o$} (B);
\end{tikzpicture}
\]
which can also be thought of as an open graph from $X$ to $Y$,
\item a 2-morphism
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$X$};
\node (B) at (1,0) {$N$};
\node (C) at (2,0) {$Y$};
\node (A') at (0,-1) {$X'$};
\node (C') at (2,-1) {$Y'$};
\node (D) at (1,-1) {$N'$};
\node (E) at (3,0) {$G \in F(N)$};
\node (F) at (3,-1) {$G' \in F(N')$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$i$} (B)
(C) edge node[above]{$o$} (B)
(A) edge node[left]{$f$} (A')
(C) edge node[right]{$g$} (C')
(C') edge node [above] {$o'$} (D)
(A') edge node [above] {$i'$} (D)
(B) edge node [left] {$h$} (D);
\end{tikzpicture}
\]
is a commuting diagram in $\Fin\Set$ together with a morphism $\tau \maps F(h)(G) \to G'$ in $F(N')$.
\end{itemize}

@@

We can also describe $\La$-circuits using decorated cospans.   There is a symmetric lax monoidal pseudofunctor $F \maps (\Fin\Set, +) \to (\Cat, \times)$ such that for any finite set $N$, the category $F(N)$ has:
\begin{itemize}
\item objects being $\La$-\define{graph structures} on $N$: that is, $\La$-graphs where the set of nodes is $N$,
\item morphisms being morphisms of $\La$-graphs that are the identity on the set of nodes.
\end{itemize}
This gives a symmetric monoidal double category $F \lCsp$, and using \cref{thm:equiv} we can show that this is equivalent, as a symmetric monoidal double category, to ${}_L \lCsp(\Graph_\La)$.

@@

There is a category $\Petri$ with Petri nets as objects, where a morphism from the Petri net 
$s, t \maps T \to \N[S]$ to the Petri net $s', t' \maps T' \to \N[S']$ is a pair of functions $f \maps T \to T', g \maps S \to S'$ such that the following diagrams commute:
	\[
	\xymatrix{ 
		T \ar[d]_f  \ar[r]^-{s} & \N[S] \ar[d]^-{\N[g]} \\	
		T' \ar[r]^-{s'} & \N[S'] 
	}
	\qquad
	\xymatrix{ 
		T \ar[d]_f  \ar[r]^-{t} & \N[S] \ar[d]^-{\N[g]} \\	
		T' \ar[r]^-{t'} & \N[S'] . 
	}
	\]
There is a functor $R \maps \Petri \to \Set$ sending any Petri net to its set of places, and this has a left adjoint $L \maps \Set \to \Petri$ sending any set $S$ to the Petri net with $S$ as its set of species and no transitions \cite[Lemma 11]{BM}.   Since both $\Set$ and $\Petri$ have finite colimits and $L$ preserves them, \cref{thm:SC} yields a symmetric monoidal double category ${}_L \lCsp(\Petri)$ in which:
\begin{itemize}
\item objects are finite sets,
\item vertical 1-morphisms are functions,
\item horizontal 1-cells as \define{open Petri nets}, which are cospans in $\Petri$ of the form:
\[
\begin{tikzpicture}[scale=1.5]
\node (D) at (-3,0) {$L(X)$};
\node (E) at (-2,0) {$P$};
\node (F) at (-1,0) {$L(Y)$};
\path[->,font=\scriptsize,>=angle 90]
(D) edge node [above] {$i$} (E)
(F) edge node [above] {$o$} (E);
\end{tikzpicture}
\]
\item 2-morphisms are maps of cospans in $\Petri$ of the form:
\[
\begin{tikzpicture}[scale=1.5]
\node (E) at (3,0) {$L(X_1)$};
\node (F) at (5,0) {$L(Y_1)$};
\node (G) at (4,0) {$P_1$};
\node (E') at (3,-1) {$L(X_2)$};
\node (F') at (5,-1) {$L(Y_2)$.};
\node (G') at (4,-1) {$P_2$};
\path[->,font=\scriptsize,>=angle 90]
(F) edge node[above]{$O_1$} (G)
(E) edge node[left]{$L(f)$} (E')
(F) edge node[right]{$L(g)$} (F')
(G) edge node[left]{$\alpha$} (G')
(E) edge node[above]{$I_1$} (G)
(E') edge node[above]{$I_2$} (G')
(F') edge node[above]{$O_2$} (G');
\end{tikzpicture}
\]
\end{itemize}

We can equivalently describe open Petri nets using decorated cospans.  This works very much like the previous examples.  There is a symmetric lax monoidal pseudofunctor $F \maps (\Fin\Set, +) \to (\Cat, \times)$ such that for any finite set $S$, the category $F(S)$ has:
\begin{itemize}
\item objects given by Petri nets whose set of places is $S$,
\item morphisms given by morphisms of Petri nets that are the identity on the set of places.
\end{itemize}
This gives a symmetric monoidal double category $F \lCsp$, and using \cref{thm:equiv} we can show that this is equivalent, as a symmetric monoidal double category, to ${}_L \lCsp(\Petri)$.

The machinery of structured cospans has been used to provide a semantics for open Petri nets \cite{BM}: a symmetric monoidal double functor from ${}_L \lCsp(\Petri)$ to  a symmetric monoidal double category of `open commutative monoidal categories'.  Presumably this double functor can equivalently be obtained using the machinery of decorated cospans, with the help of \cref{thm:functoriality}.  However, it should be clear by now that in cases where either structured or decorated cospans can be used, structured cospans are simpler.   We next turn to an example where decorated cospans are necessary.

@@

Let $\Dbl$ denote the 2-category of double categories, double functors and double transformations. One can check that $\Dbl$ has finite products, and in any 2-category with finite products we can define a `pseudomonoid', which is a categorified analogue of a monoid \cite{DS}. For example, a pseudomonoid in $\Cat$ is a monoidal category.   We can also define symmetric pseudomonoids, which in $\Cat$ are symmetric monoidal categories.
